\large How resource abundance and stochasticity affect animals' spatial needs

### 2022-11-21 writing/manuscript.Rmd
added to the discussion and some refs

Extreme events [@logares_black_2012; @anderson_black-swan_2017] such as heat domes, blizzards, fires, disease outbreaks, and oil spills are important events which are hard to predict (i.e., highly stochastic) and can cause great population instability by causing mass die-offs that drastically reduce the size and genetic diversity of a population [e.g., @matkin_ongoing_2008; @murray_pathogens_2006; @widmer_effects_2004; @reed_frequency_2003; @hollings_trophic_2014; @ameca_y_juarez_natural_2012; @berger_climate_2018]. Managing the intensity of and frequency of extreme events (e.g., controlled burns) can increase the stability of a habitat while promoting habitat diversity and adaptations to the event [@staver_tree_2011; @tingley_pyrodiversity_2016; @nimmo_animal_2019].

@@

\noindent This paper presents a new unifying framework for estimating the effects of resource abundance and stochasticity on animals' spatial needs. The hypotheses and methods we present allow researchers to generalize the effects resource abundance and stochasticity have on animals' spatial needs by building upon previous work rather than stating their results as *de novo* findings. Instead, one can test whether members of the same population or species behave similarly, and whether all members of the group have a common functional response [@ref]. Individuals and populations that were once thought to have markedly different behaviors and spatial needs may instead have a single common functional response but different levels of $\e(R)$ and $\var(R)$ [@examples].

@@

@liow_lower_2009: *Lower Extinction Risk in Sleep-or-Hide Mammals*

\noindent The spatiotemporal scales over which an event occurs are a main determinant of whether an animal will be able to predict the event and how it will respond to it (or fail to do so). Events which occur at the scale of the organism [i.e. approximately one or more times per lifetime and within the animal's spatial range; see @frankham_importance_2004] are more likely to be perceived as a threat or boon worth preparing for [e.g., elephants: @foley_severe_2008; bison and mule deer: @geremia_migrating_2019]. Smaller and short-lived organisms (e.g., mice) are more likely to be be severely impacted by stochastic events than larger ones (e.g. elephants), since larger organisms can have bigger energy reserves [@lindstedt_seasonality_1985], can move longer distances over short periods of time [@hirt_general_2017], and tend to have longer lifespans, generation times, and developmental periods [@brown_toward_2004] that allow them to develop or memory about the frequency and severity of such events [@foley_severe_2008; @polansky_elucidating_2015]. However, the short generation time and high reproductive rate of smaller *r-selected* species [@pianka_r-_1970; @brown_toward_2004] may also promote traits that increase survival following frequent extreme events by increasing population survival rates and genetic diversity through the production of numerous offspring [@examples]. Frequent events may result in changes in animal behavior and memory [framework: @riotte-lambert_environmental_2020; elephants: @foley_severe_2008; bison and mule deer: @geremia_migrating_2019; elk: @falcon-cortes_hierarchical_2021], while events that occur over larger spatiotemporal scales are more likely to promote slower changes through adaptation, natural selection, and evolution [@grant_evolution_2017; @logares_black_2012; @anderson_black-swan_2017; @gienapp_climate_2008], including the developing of memory [framework: @riotte-lambert_environmental_2020; @fagan_spatial_2013; roe deer: @ranc_memory_2022; elk: @falcon-cortes_hierarchical_2021; blue whales: @abrahms_memory_2019; elephants: @polansky_elucidating_2015].

* range shifts can help species overcome changing conditions, but they can also damage species that are being encroached upon, such as species that either fail to adapt (i.e., do not shift, too) or species that **cannot** shift

Adaptability and responsiveness to change are strong determinants to whether an animal survives in a changing environment [@riotte-lambert_environmental_2020]. An animal's ability to recognize cues that precede extreme events or periods of scarcity or abundance can have appreciable effects on the animal's chance of survival and reproduction. For example, ungulates' ability to time their yearly migrations on cues such as available forage biomass, hunting pressure, snow fall, and snow melt allow them to adapt to changes in both $\e(R)$ and $\var(R)$, while their spatial memory [@falcon-cortes_hierarchical_2021; @ranc_memory_2022] allows them to optimize their movement by reducing the amount of time spent exploring. Similarly, the ability of trans-Saharian migrant birds to adapt their migration timing based on environmental cues allows them to account for differences $\e(R)$ between years [@jonzen_rapid_2006]. However, relying strongly on short-term cues rather than depending on memory can also cause animals to fail to find crucial resources [@polansky_elucidating_2015; @foley_severe_2008] or become trapped in sub-optimal locations if long-term conditions and cycles change unexpectedly [e.g., blue whales: @abrahms_memory_2019; bison: @merkle_spatial_2019].

@@

@rickbeil_plasticity_2019:
- Many species use seasonal migration to track emerging resources (Fryxell, Greever, & Sinclair, 1988), reduce predation risk (Hebblewhite & Merrill, 2009), or avoid disease and parasites (Altizer, Bartel, & Han, 2011).
- Migrations to track resources are often timed to coincide with peak plant and insect emergence (Aikens et al., 2017; Jonzén et al., 2006; Merkle et al., 2016)
- there is growing evidence that some [bird] populations are adjusting their behavior, suggesting that migration, specifically migration timing, may be more plastic than previously expected (Jonzén, Hedenstrom, & Lundberg, 2007; Jonzén et al., 2006; Lameris et al., 2018) and highlighting a potential measure of resilience in the face of climate change
-->

@@

The ability of a specialist or obligate symbiont to shift its home range or adapt will likely depend strongly on its food source's or associate's ability to move or adapt, too.

>>>>

Extreme events [@logares_black_2012; @anderson_black-swan_2017] such as heat domes, hurricanes [@widmer_effects_2004]; severe storms [@berger_climate_2018], fires [@ruthrof_how_2016], new diseases [@murray_pathogens_2006; @hollings_trophic_2014; @hoyt_ecology_2021], and anthropogenic die-offs [@matkin_ongoing_2008] are important events which are hard to predict (i.e., highly stochastic) and can cause great population instability by drastically reducing the size and genetic diversity of a population [@reed_frequency_2003; @ameca_y_juarez_natural_2012]. Managing the intensity of and frequency of extreme events [e.g., @aponte_forest_2016] can increase the stability of a habitat while promoting habitat diversity and adaptations to the event [@staver_tree_2011; @tingley_pyrodiversity_2016; @nimmo_animal_2019].

@@

\noindent This paper presents a new unifying framework for estimating the effects of resource abundance and stochasticity on animals' spatial needs. The hypotheses and methods we present allow researchers to generalize the effects resource abundance and stochasticity have on animals' spatial needs by building upon previous work rather than stating their results as *de novo* findings. Instead, one can test whether members of the same population or species behave similarly, and whether all members of the group have a common functional response using hierarchical models [@pedersen_hierarchical_2019]. Individuals and populations that were once thought to have markedly different behaviors and spatial needs may instead have a single common functional response but different levels of $\e(R)$ and $\var(R)$ [@examples].

@@

\noindent The spatiotemporal scales over which an event occurs are a main determinant of whether an animal will be able to predict the event and how it will respond to it (or fail to do so). Events which occur at the scale of the organism [i.e. approximately one or more times per lifetime and within the animal's spatial range; see @frankham_importance_2004] are more likely to be perceived as a threat or boon worth preparing for [e.g., elephants: @foley_severe_2008; bison and mule deer: @geremia_migrating_2019]. Smaller and short-lived organisms (e.g., mice) are more likely to be be severely impacted by stochastic events than larger ones (e.g. elephants), since larger organisms can have bigger energy reserves [@lindstedt_seasonality_1985], can move longer distances over short periods of time [@hirt_general_2017], and tend to have longer lifespans, generation times, and developmental periods [@brown_toward_2004] that allow them to develop or memory about the frequency and severity of such events [elephants: @foley_severe_2008; @polansky_elucidating_2015]. However, the short generation time and high reproductive rate of smaller *r-selected* species [@pianka_r-_1970; @brown_toward_2004] may also promote traits that increase survival following frequent extreme events by increasing population survival rates and genetic diversity through the production of numerous offspring [@examples]. Frequent and repetitive events may result in changes in animal behavior and memory [framework: @riotte-lambert_environmental_2020; elephants: @foley_severe_2008; bison and mule deer: @geremia_migrating_2019; elk: @falcon-cortes_hierarchical_2021], while events that occur over larger spatiotemporal scales are more likely to promote slower changes through adaptation, natural selection, and evolution [@grant_evolution_2017; @logares_black_2012; @anderson_black-swan_2017; @gienapp_climate_2008].

Events occur which multiple times at relatively predictable intervals during an animal's lifespan and over multiple generations, such as the alternating of seasons, allow it to anticipate times of scarcity and prepare for them by, for instance, increasing its fat storage [marmots: @armitage_energetics_2003; bears: @nespolo_why_2022] or caching resources [@post_caching_2006]. Of course, an animal with a life span shorter than a year will be unable to perceive the cyclical alternating of the seasons, and an animal with a life span of a few weeks or days may not even experience substantial changes in weather and temperature other than lower temperatures at night than during the day. When an animal's expected lifespan is at least one or more years, preparing for predictable times of abundance and scarcity becomes an advantage (and often a necessity). In such cases, animals can increase their chances of survival via physiological and behavioral adaptations, including torpor [marmots: @inouye_climate_2000; ground squirrels: @goldberg_hibernation_2021]; and memory [framework: @riotte-lambert_environmental_2020; @fagan_spatial_2013; roe deer: @ranc_memory_2022; caribou: @cavedon_genomic_2022; elk: @falcon-cortes_hierarchical_2021; elephants: @polansky_elucidating_2015], which is required for migration [butterflies: @stefanescu_timing_2021; bison and mule deer: @geremia_migrating_2019; mule deer: @merkle_spatial_2019; blue whales: @abrahms_memory_2019; elk: @rickbeil_plasticity_2019; @middleton_green-wave_2018]. However, relying on environmental cues rather than on memory alone [ungulates: @severson_spring_2021; @rickbeil_plasticity_2019; @buntgen_elevational_2017; birds: @jonzen_rapid_2006], can prevent animals from becoming trapped in sub-optimal conditions [@abrahms_memory_2019].

Adaptability and responsiveness to change are strong determinants to whether an animal survives in a changing environment [@riotte-lambert_environmental_2020]. An animal's ability to recognize cues that precede extreme events or periods of scarcity or abundance can have appreciable effects on the animal's chance of survival and reproduction. For example, yellow-bellied marmots (**) []

Likewise, ungulates' ability to time their yearly migrations on cues such as available forage biomass, hunting pressure, snow fall, and snow melt allow them to adapt to changes in both $\e(R)$ and $\var(R)$, while their spatial memory [@falcon-cortes_hierarchical_2021; @ranc_memory_2022] allows them to optimize their large-scale movement by reducing the amount of time spent exploring. Similarly, the ability of trans-Saharan migrant birds to adapt their migration timing based on environmental cues allows them to account for differences $\e(R)$ between years [@jonzen_rapid_2006]. However, relying strongly on short-term cues rather than depending on memory can also cause animals to fail to find crucial resources [@polansky_elucidating_2015; @foley_severe_2008].

Recent changes in climate during the last few decades have caused some species to shift their ranges [terrestrial animals: @chen_rapid_2011; ungulates: @severson_spring_2021; @buntgen_elevational_2017], including migratory animals [@examples]. While range shifts can help species overcome changing conditions, they can still result in phenological mismatches [@lameris_arctic_2018], and they also add pressure to the species that already live in the habitat, such as species that either fail to adapt (i.e., do not also shift) or species that are unable shift. Specialists and obligate symbionts are less likely to shift since their range is determined by their food sources' or associates' ability to move and adapt, too. Consequently, the arrival of new species may exacerbate the stressors such animals are already facing.

@@

@rickbeil_plasticity_2019:
- Many species use seasonal migration to track emerging resources (Fryxell, Greever, & Sinclair, 1988), reduce predation risk (Hebblewhite & Merrill, 2009), or avoid disease and parasites (Altizer, Bartel, & Han, 2011).
- Migrations to track resources are often timed to coincide with peak plant and insect emergence (Aikens et al., 2017; Jonzén et al., 2006)
- there is growing evidence that some [bird] populations are adjusting their behavior, suggesting that migration, specifically migration timing, may be more plastic than previously expected (Jonzén, Hedenstrom, & Lundberg, 2007; Jonzén et al., 2006) and highlighting a potential measure of resilience in the face of climate change
-->

@@
